Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-15 LICS/sections/appendix.tex
wip

\section{DPO rewriting for arbitrary signatures}
\label{sec:dpo-fix}
Here we will discuss how one can formulate DPO rewrite rules where the left-hand of right-hand sides contain sub-hypergraphs with no inputs and outputs at the outermost level.
First we will illustrate where the restriction in the original formulation of the DPO rewriting as per the Definition~\ref{def:convex_dpo} comes from via several examples: one for Cartesian equations (which are important to express e-graphs as e-hypergraphs) and two for the structural rules we need for $\catname{MEHypI}_{\Sigma} / \mathcal{S}$ to be a category enriched over $\catname{SLat}$.
Consider a rewrite rule induced by the Cartesian structure $a;! = id_{I}$ for $a : 0 \to 1$ which is rendered string diagrammatically as 
\[
\tikzfig{../figures/appendix/cartesian_rewrite_string}
\]
which becomes the following extended cospan of e-hypergraphs
\[
\tikzfig{../figures/appendix/cartesian_rewrite_hypergraph}
\]

@@

\begin{definition}[Hashcons invariant]
The hashcons $H$ must map all canonical e-nodes to their e-class ids:
\[
\text{e-node } n \in M[a] \text{ iff } H[\text{canonicalize}(n)] = \text{find}(a)
\]
\end{definition}

The pseudocode below shows the translation from a \textit{canonical} (the one for which the above invariant holds).
For the following e-graph with e-classes shown as dashed boxes it produces the following cospan of e-hypergraphs.
\texttt{connect} and \texttt{mkEdge} are responsible for making things shared using $\triangle$.

@@

\begin{algorithm}
    \caption{E-graph to e-hypergraph translation}
    \begin{algorithmic}[1]
    \Function{e-class-to-HEdge}{e\_class\_id}
    \State e\_nodes $\gets$ [e\_node$_{1}$ $\ldots$ e\_node$_{n}$] \textbf{in} H.keys() \textbf{where} H[e\_node$_{i}$] == M[e\_class\_id]
    % \State \textbf{for} [$\text{e_node}_{1}, \ldots, \text{e_node}_{n}$] \textbf{in} H \textbf{where} H[$\text{e_node}_{i}$] == M[e-class-id] \texbf{do}
    \ForAll {e\_node$_{i}$ in e\_nodes}
    \State edge $\gets$ mkEdge(e\_class\_id)
    \Comment {Create a hierarchical edge that represents an e-class}
    \State \textbf{match} e\_node$_{i}$ \textbf{with}
    \State \qquad \textbf{case} f($e_1 \ldots e_{n}$) :
    \State \qquad \qquad \textbf{if} is\_singleton(e\_class\_id) \textbf{then}
    % \If{is\_singleton(e\_class\_id)}
        \State \qquad\qquad \hspace{2em} edge $\gets$ mkEdge(f)
        \Comment {Create an edge that represents the generator f $: k \to 1$}
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
        \State \qquad \qquad \hspace{2em} \Return edge
    % \Else
    \State \qquad \qquad \textbf{else}
    \State \qquad \qquad \hspace{2em} graph $\gets$ discard([e\_node$_1$$\ldots$e\_node$_{i-1}$]) $\cup$ mkEdge(f) $\cup$ discard([e\_node$_{i+1}$ $\ldots$ e\_node$_{n}$])
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.add(graph)
        \Comment {Add \texttt{graph} as a consistent component into hierarchical edge \texttt{edge}}
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
    % \EndIf
    \EndFor
    \State \Return edge
    % \Comment {There will be a single e-node in a singleton e-class}
    % \ForAll {e-node in H \textbf{where} H[e-node] == M[e-class-id]}
    %     \State \textbf{match} e-node \textbf{with}
    %     \State \quad \textbf{case} \texttt{f}: \Return (mkEdge(f))
    %     \State \quad \textbf{case} \texttt{f}($e_1$, $\ldots$, $e_{n}$) :
    %     % this is nested
    %         \State \qquad\qquad edge $\gets$ mkEdge(f)
    %         \State \qquad\qquad edge.connect([e-class-to-HEdge(e) | e $\gets$ [$e_1$, \ldots, $e_{n}$]])
    %         \State \qquad\qquad \Return edge
    % \EndFor
    % \Else
    % \State edge $\gets$ mkEdge(M[e-class-id])
    % \ForAll {e-node in H \textbf{where} H[e-node] == M[e-class-id]}
    %     \State \textbf{match} e-node \textbf{with}
    %     \State \quad \textbf{case} \texttt{f}: edge.add(f)
    %     \State \quad \textbf{case} \texttt{f}($e_1$, $\ldots$, $e_{n}$): edge.add(f)
    % \EndFor
    % \EndIf
    \EndFunction
    \State
    \State graph $\gets \varnothing$ 
    \ForAll{e\_class\_id in H.values()}
    \State graph $\gets$ graph $\cup$ \Call {e-class-to-HEdge}{e\_class\_id})

    \EndFor
    % \Function{matchPattern}{shape}
    %     \State \textbf{match} shape \textbf{with}
    %     \State \quad \textbf{case} \texttt{Circle(radius)}:
    %     \State \quad \quad \Return $\pi \cdot \texttt{radius}^2$
    %     \State \quad \textbf{case} \texttt{Rectangle(width, height)}:
    %     \State \quad \quad \Return \texttt{width} $\times$ \texttt{height}
    %     \State \quad \textbf{case} \texttt{Triangle(base, height)}:
    %     \State \quad \quad \Return $\frac{1}{2} \times \texttt{base} \times \texttt{height}$
    %     \State \quad \textbf{case} \texttt{Square(side)}:
    %     \State \quad \quad \Return \texttt{side}$^2$
    %     \State \quad \textbf{case} \texttt{\_}:
    %     \State \quad \quad \Return ``Unknown shape''
    % \EndFunction
    \end{algorithmic}
\end{algorithm}

\begin{lemma} Two e-classes $e_1$ and $e_2$ are merged if and only if there are rewrite rules that turn every term representable by $e_1$ into $e_2$.

\begin{lemma}
Every hierarchical edge can be decomposed as follows.

\[
\tikzfig{../figures/appendix/hyperedge_decomposition}
\]
\end{lemma}

Then, given e-classes $e_1$ and $e_2$ for sub-hypergraphs $C$ and $D$ respectively, we introduce e-nodes into the e-graph $A(e_1)$ and $B(e_2)$ that reside in e-class $e_3$.

>>>>

\section{DPO rewriting for arbitrary signatures}
\label{sec:dpo-fix}
Here we will discuss how one can formulate DPO rewrite rules where the left-hand of right-hand sides contain sub-hypergraphs with no inputs and outputs at the outermost level.
First we will illustrate where the restriction in the original formulation of the DPO rewriting as per the Definition~\ref{def:convex_dpo} comes from via several examples: one for Cartesian equations (which are important to express e-graphs as e-hypergraphs) and two for the structural rules we need for $\WellTypedMdaEcospans / \mathcal{S}$ to be a category enriched over $\catname{SLat}$.
Consider a rewrite rule induced by the Cartesian structure $a;! = id_{I}$ for $a : 0 \to 1$ which is rendered string diagrammatically as 
\[
\tikzfig{../figures/appendix/cartesian_rewrite_string}
\]
which becomes the following extended cospan of e-hypergraphs
\[
\tikzfig{../figures/appendix/cartesian_rewrite_hypergraph}
\]

@@

\begin{definition}[Congruence invariant~\cite{EggPaper}]
Two e-nodes $f(a_{1}, \ldots, a_{n})$ and $f_{b_{1}, \ldots, b_{n}}$ are \textit{congruent}, if $find(a_{i}) = find(b_{i})$ for all $i$.
The equivalence relation on e-nodes must be closed under this congruence, that is, two congruent e-nodes must reside in the same e-class.
\end{definition}

\begin{definition}[Hashcons invariant~\cite{EggPaper}]
The hashcons $H$ must map all canonical e-nodes to their e-class ids:
\[
\text{e-node } n \in M[a] \text{ iff } H[\text{canonicalize}(n)] = \text{find}(a)
\]
\end{definition}

\begin{definition}[~\cite{EggPaper}]
A rewriting over an e-graph is given by composition of the following operations
\begin{itemize}
    \item \textit{add} takes an e-node $n$ and:
    \begin{itemize}
        \item if $n$ is present in $H$, return $H[n]$;
        \item else create e-class id $a$ and set $H[n] = a$ such that $M[a] = \{n\}$.
    \end{itemize}
    \item \textit{merge} given two e-class ids $a$ and $b$ unions them in the union find $U$ such that $M[a] = M[b] = M[a] \cup M[b]$.
\end{itemize}
\end{definition}
Then, given a rewrite rule $l \to r$, as, for example, $x * 2 \to x \ll 1$ in Figure~\ref{fig:e-graph-example}, applying it to an e-graph amounts several steps.
First, finding a tuple $(\sigma, c)$ where e-class id $c$ contains $\sigma(l)$ where sigma is the mapping from variables to e-class ids, for example, $c = H[a * 2]$ and $\sigma(l) = (a * 2)$.
Then, adding $\sigma(r)$ to the e-graph and merging $c$ with $add(\sigma(r))$.
The latter step may require additional merges to maintain the invariant.
If two terms $f(a,b)$ and $f(a,c)$ reside in different e-classes, merging e-classes for $a$ and $b$ would require merging e-classes for $f(a,b)$ and $f(a,c)$ as they become congruent as congruent e-nodes must reside in the same e-class.
The process of maintaining the invariants by additional merging is called \textit{upward merging}~\cite{EggPaper}.
We will denote the process of obtaining e-graph $e_2$ from e-graph $e_1$ after applying a rule $l \to r$ (including upward merging) as $e_1 \leadsto e_2$.

\begin{definition}
The pseudocode in Algorithm~\ref{alg:translation} defines the translation function $\llbracket \rrbracket : \catname{E}\text{-}\catname{graph} \to \WellTypedMdaEcospans$ that turns a \textit{canonical} (the one for which the above invariants hold) to a cospan of e-hypergraphs.
\end{definition}

For the following e-graph with e-classes shown as dashed boxes it produces the following cospan of e-hypergraphs.
\texttt{connect} and \texttt{mkEdge} are responsible for making things shared using $\triangle$ and discarding unused inputs via $!$.

@@

\begin{algorithm}
    \caption{E-graph to e-hypergraph translation}
    \label{alg:translation}
    \begin{algorithmic}[1]
    \Function{e-class-to-HEdge}{e\_class\_id}
    \State e\_nodes $\gets$ [e\_node$_{1}$ $\ldots$ e\_node$_{n}$] \textbf{in} H.keys() \textbf{where} H[e\_node$_{i}$] == M[e\_class\_id]
    % \State \textbf{for} [$\text{e_node}_{1}, \ldots, \text{e_node}_{n}$] \textbf{in} H \textbf{where} H[$\text{e_node}_{i}$] == M[e-class-id] \texbf{do}
    \ForAll {e\_node$_{i}$ in e\_nodes}
    \State edge $\gets$ mkEdge(e\_class\_id)
    \Comment {Create a hierarchical edge that represents an e-class}
    \State \textbf{match} e\_node$_{i}$ \textbf{with}
    \State \qquad \textbf{case} f($e_1 \ldots e_{n}$) :
    \State \qquad \qquad \textbf{if} is\_singleton(e\_class\_id) \textbf{then}
    % \If{is\_singleton(e\_class\_id)}
        \State \qquad\qquad \hspace{2em} edge $\gets$ mkEdge(f)
        \Comment {Create an edge that represents the generator f $: k \to 1$}
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
        \State \qquad \qquad \hspace{2em} \Return edge
    % \Else
    \State \qquad \qquad \textbf{else}
    \State \qquad \qquad \hspace{2em} graph $\gets$ discard([e\_node$_1$$\ldots$e\_node$_{i-1}$]) $\cup$ mkEdge(f) $\cup$ discard([e\_node$_{i+1}$ $\ldots$ e\_node$_{n}$])
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.add(graph)
        \Comment {Add \texttt{graph} as a consistent component into hierarchical edge \texttt{edge}}
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
    % \EndIf
    \EndFor
    \State \Return edge
    \EndFunction
    \State
    \State graph $\gets \varnothing$ 
    \ForAll{e\_class\_id in H.values()}
    \State graph $\gets$ graph $\cup$ \Call {e-class-to-HEdge}{e\_class\_id})

    \EndFor
    \end{algorithmic}
\end{algorithm}

\begin{lemma} 
    Two e-classes $e_1$ and $e_2$ are merged if and only if there are rewrite rules that turn every term representable by $e_1$ into $e_2$.
\end{lemma}

\begin{lemma}
Given a set of equations $\mathcal{E}$ between $\Sigma$-terms that induces a set of rewrite rules $l \to r$ and $r \to l$ for every $l = r \in \mathcal{E}$, $e_1 \leadsto e_2$ if and only if $\llbracket e_1 \rrbracket \Rrightarrow{}^{*} \llbracket e_2 \llbracket$.
\end{lemma}
\begin{proof}

We will prove this by example showing how an addition of an e-node followed by a series of merges can be expressed as a series of DPO rewrites.
Consider an example in Figure~\ref{fig:e-graph-rewrite}.
It shows the rewrite $b \to c$ applied to e-graph that initially encodes a single term $f(a,b) + f(a,c)$.
The first step is $\textit{merge}(b, \textit{add}(c))$ which unions e-classes for $b$ and $c$ by making the latter be the parent of the former, \textit{i.e.}, \text{find}$(b) = c$.
Then it recanonicalizes the e-node for $f(a,b)$ so that it points to the canonical e-class id for $b$.
This makes it possible to further merge two e-classes as they contain congruent e-nodes.
The interpretation of these steps as DPO rewrites is given in Figure~\ref{fig:e-string-rewrite}.
Subfigure $(a)$ is the direct translation of the initial e-graph by using $\llbracket \rrbracket$.
Then we utilize the idempotence twice as $b = b + b$ and $c = c + c$ (subfigure (b)) and use the fact that our rewrite rules are symmetric as they are induced by the set of equations and apply $b \to c$ and $c \to b$ (subfigure (c)) and then use the rules for sharing from the Cartesian structure (subfigures (d)-(e)).
The reverse direction follows by finding normal forms for $\llbracket e_{1} \rrbracket$ and $\llbracket e_{2} \rrbracket$.
\end{proof}

\begin{remark}

Technically, rewrite rules formed by $\Sigma$-terms do not contain any variables as they are not a part of the signature while the rewrite rules for a given e-graph do contain variables as, for example, $x * 2 \to x \ll 1$.
This is however not a limitation, as variables are merely interfaces and we can express the above-mentioned rule as a pair of $\Sigma$-terms $*;id \otimes 2 \to \ll ; id \otimes 1$.
\end{remark}

\[
\adjustbox{width=\linewidth}{
    \tikzfig{../figures/appendix/e-graph-rewrite-example}
}
\]
\caption{E-graph rewrite example.}
\label{fig:e-graph-rewrite}
\end{figure}

    \[
        \adjustbox{width=\linewidth}{
            \tikzfig{../figures/appendix/e-string-rewrite-example}
        }
        \]
        \caption{String diagram rewrite example.}
        \label{fig:e-string-rewrite}
\end{figure}

% \begin{lemma}
% Every hierarchical edge can be decomposed as follows.

% \[
% \tikzfig{../figures/appendix/hyperedge_decomposition}
% \]
% \end{lemma}

% Then, given e-classes $e_1$ and $e_2$ for sub-hypergraphs $C$ and $D$ respectively, we introduce e-nodes into the e-graph $A(e_1)$ and $B(e_2)$ that reside in e-class $e_3$.
